NMRlipids Databank makes data-driven analysis of biomembrane properties accessible for all

### 2023-04-17 main.tex
Updates from Overleaf

Here we present the NMRlipids Databank---a community-driven, open-for-all database featuring programmatic access to atom-resolution MD simulations of lipid bilayers. The programmatic access enables users to apply their data-driven approaches on the Databank, thus facilitating the creation of new tools for (and by) researchers in a wide range of fields covering academia and industry, from cell membrane biology and lipid nanoparticle formulations to computational chemistry and machine learning. As two usage examples of what is already possible, we demonstrate here (i) how data-driven analysis of water anisotropic diffusion in all the membrane systems available in the Databank can extend the scope of MD simulations to magnetic resonance imaging (MRI) and pharmacokinetics and (ii) how the Databank allows its users to analyse rare phenomena that are beyond the scope of standard MD simulation investigations. Wider adaptation of the NMRlipids Databank will open even more possibilities. Furthermore, the Databank performs automatic quality evaluation of membrane simulations, which facilitates the selection of best-performing models for each given application and accelerates the development of simulation parameters and methodology.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\linewidth]{Figures/overlay3.pdf}
    \caption{A) Schematic presentation of the overlay structure used in the NMRlipids Databank. A more detailed structure of the \textit{Databank layer} is shown in Fig.~\ref{DatabankStructure} in the SI.
    %b) Distribution of the lengths of the trajectories, total number of trajectories and total length of the simulations in the NMRlipids databank.
    B) Distribution of lipids present in the trajectories of the Databank. `Others' lists lipids occurring in five or fewer simulations. 
    C) Distribution of force fields in the simulations in the Databank. References for each force field are given in Table~\ref{tab:ForceFields}.
    D) Flowchart for performing an analysis of properties through all MD simulations in the NMRlipids Databank using the API.
    E) Flowchart for accessing results calculated from the NMRlipids Databank and stored to the \textit{Application layer}.
    F) Currently available binary mixtures in the NMRlipids Databank. 
    }
    \label{fig:overlay}
\end{figure}

@@

\begin{figure}[!t]
    \centering
    \includegraphics[width=\linewidth]{Figures/quality5.pdf}
    \caption{ A) Area per lipid of POPC and POPE lipid bilayers predicted by different force fields at 310\,K in simulations that are available in the NMRlipids Databank. The data points from the best-performing simulations, based on rankings in panels B and C, are surrounded by black circles. 
    B) Best POPC simulations ranked based on the \textit{sn}-1 acyl chain order parameter quality.
    C) Best POPE simulations ranked based on the \textit{sn}-1 acyl chain order parameter quality. 
    D)--F) Direct comparison against experimental (NMR order parameters and X-ray scattering ) data exemplified for a simulation with the best overall order parameter quality (D), the best quality for POPE lipid (E), and the headgroup quality for POPE (F).
    G) Scatter plots and Pearson correlation coefficients, $r$, for the membrane area per lipid, thickness, first minimum of X-ray scattering form factor and average order parameter of the \textit{sn}-1 acyl chain extracted from the NMRlipids Databank. All correlation coefficients have p-value below 0.001 as indicated by~$^{***}$. For more correlations see Fig.~\ref{fig:QualityCorrelationsSI}.
    }
    \label{fig:quality}
\end{figure}

@@

\begin{figure}[htb]
    \centering
    \includegraphics[width=88mm]{Figures/CholFlipFlops.pdf}
    \caption{\textbf{A} Illustration of cholesterol flip-flop.  
      \textbf{B--D} Cholesterol flip-flops analyzed from the Databank as a function of membrane thickness, area per lipid, and acyl chain order. Values from simulations with non-zero flip-flop rates are shown with blue dots. Histogrammed values are shown with black dots.
    }
    \label{fig:flip-flops}
\end{figure}

@@

\begin{figure}[tb]
    \centering
    \includegraphics[width=\linewidth]{Figures/permeation2.pdf}
    \caption{\textbf{A} Water diffusion, $D_\perp$, and permeability, $P$, through membranes, and lateral diffusion along the membrane, $D_\parallel$, illustrated in a multilamellar stack of lipid bilayers. 
    \textbf{B--E} Water permeation through membranes analyzed from the databank as a function of temperature, thickness, area per lipid, and acyl chain order. Values from simulations with non-zero permeation values are shown with blue dots. Histogrammed values are shown with black dots. Inset in B) shows the Arrhenius plot of permeation ($\ln(P)$ vs. $1/T$) that gives 17\,$\pm$\,4\,$k_\mathrm{B}T$ for the average activation energy for water permeation through lipid bilayer. Insets in C) and D) show the region where the dependence could be considered approximately linear.
    \textbf{F} Lateral diffusion of water as a function of hydration level. Experimental points for DMPC bilayers at 313\,K at different hydration levels are shown~\cite{rudakova04}.
    \textbf{G--H} Diffusion anistoropy of water as a function of thickness and area per lipid. }
    \label{fig:permeability}
\end{figure}

To this end, we first calculated the water permeability through membranes from all simulations in the NMRlipids Databank using the general workflow depicted in Fig.~\ref{fig:overlay}D. The resulting non-zero values range between 0.3--322\,\textmu{}m/s with the mean of 14\,\textmu{}m/s and median of 8\,\textmu{}m/s. These values agree with the previously reported simulation results~\cite{venable19,camilo2022}, but are on average slightly larger than experimental values reported for PC-lipids in the liquid crystalline phase,~0.19--0.33\,\textmu{}m/s~\cite{jansen95}. Using the workflow depicted in Fig.~\ref{fig:overlay}E, we then plotted the observed permeabilities and their histogrammed values in Figs.~\ref{fig:permeability}B--E as a function of temperature, membrane thickness, area per lipid, and acyl chain order. As expected, the permeability increases with the temperature, giving an average energy barrier of 17\,$\pm$\,4\,$k_\mathrm{B}T$ for the water permeation from the Arrhenius plot in Fig.~\ref{fig:permeability}B. On the other hand, the permeability of water decreases on average when membranes become more packed, \textit{i.e.}, with decreasing area per lipid and increasing thickness and acyl chain order (Figs.~\ref{fig:permeability} C--E). Permeation of water through bilayers depends on membrane properties also according to previous studies, but there is no established consensus on whether the area per lipid~\cite{nagle08} or bilayer thickness~\cite{frallicciardi22} is the main parameter determining the permeability. 
%is previously reported from simulations and experiments for some systems, but not for all~\cite{mathai07,venable19,frallicciardi22}. 
Our analysis over the NMRlipids databank, containing significantly more data than what was available in previous studies, suggest non-linear dependencies on both of these parameters, yet the linear correlation would be a good approximation for thicknesses above $\sim$\,3.9\,nm and area per lipids below $\sim$69\,\AA{}$^2$ (insets in Figs.~\ref{fig:permeability} C--D).
%within the range of experimentally reported values  permeability which is 1-2 orders of magnitude slower than osmotic permeability \cite{jansen95,??}. 
Clear dependencies of permeability on charged lipids, cholesterol, POPE, or hydration level were not observed in Fig.~\ref{fig:permeationSI} in the supplementary information.
%
%%Several models on how water dynamics depends on membrane properties have been proposed~\cite{mathai07,nitsche13,nitsche16,shinoda16,venable19,frallicciardi22}.

To analyze how water diffusion anisotropy depends on membrane properties in a multi-lamellar lipid bilayer system, we calculated the water diffusion parallel to the membrane surface from all simulations in the NMRlipids databank using the workflow depicted in Fig.~\ref{fig:overlay}D. The parallel diffusion coefficient of water, $D_\parallel$, decreases with reduced hydration and increases with the temperature, but dependencies on membrane area per lipid, thickness, or fraction of charged lipids were not observed in Figs.~\ref{fig:permeability} and~\ref{fig:diffusionSI} when the workflow depicted in Fig.~\ref{fig:overlay}E was used. Simulation results are close to the experimental values with low hydration levels in Fig.~\ref{fig:permeability}F, but increase to approximately two times higher than the experimental value for bulk water diffusion value (3.1$\cdot$\,10$^{-9}$\,m$^2$/s at 313\,K~\cite{khakimov08}) with high hydration levels. This is not surprising as the most common water model used in membrane simulations, TIP3P, overestimates the bulk water diffusion~\cite{pathirannahalage21}. To estimate the diffusion anisotropy of water, $D_\perp/D_\parallel$, in multilamellar membrane system, the permeability coefficients of water through membranes were translated to perpendicular diffusion coefficients, $D_\perp$, using the Tanner equation~\cite{tanner78,wasterby02}. The resulting perpendicular diffusion coefficients are approximately five orders of magnitude smaller than lateral diffusion coefficients of water (Figs.~\ref{fig:permeability} G and H), which is at the upper limit of the anisotropy estimated from the experimental data~\cite{nitsche19}. Significant increase in the diffusion anisotropy with membrane packing is observed, as $D_\perp/D_\parallel$ drifts away from one with decreasing area per lipid and increasing thickness in Figs.~\ref{fig:permeability} G and H. This follows from decreasing water permeability with membrane packing (Figs.~\ref{fig:permeability} C and D), while lateral diffusion remains approximately constant (Figs.~\ref{fig:diffusionSI} A and C).

In summary, our results suggest that the bilayer packing has a substantial effect on anisotropic water diffusion in multi-membrane lipid systems. Several folds larger anisotropy in membranes with higher lateral density is expected to play a role in pharmacokinetic models not only for water but also for other hydrophilic molecules~\cite{nitsche19}. Furthermore, the enhanced understanding of this anisotropy may help in developing new diffusion tensor based MRI imaging methods where signals originate from anisotropic diffusion of water in biological material~\cite{topgaard20}.

@@

The focus of biomolecular simulations is moving from studies of individual molecules to larger complexes and even whole cells and organelles~\cite{johnson15,thornburg22,gupta22}. Simultaneously, machine learning -based models predicting behaviour of biomolecules and automatic approaches to parametrize models are emerging~\cite{jumper21,antila22b}. The resources delivered by the NMRlipids databank will support the development in both of these directions. Automatic quality evaluation and ranking of simulations against experimental data enables the selection of best simulations for specific applications without laborious manual force field evaluation. This also streamlines automatic parametrization procedures for atomistic and coarse grained simulations. Such practises fostering the accuracy of simulations are becoming increasingly important due the accumulation of small errors when complexity and size of simulated systems are increasing.

The open programmatic access to unprecedented amount of MD simulation data \textit{via} NMRlipids databank-API enables wide range of novel data-driven analyses and machine learning applications. The analyses of cholesterol flip--flop events (Fig.~\ref{fig:flip-flops}) and water permeation through membranes (Fig.~\ref{fig:permeability}) demonstrate how large amount of accessible simulation data in terms of quantity (\textit{e.g.}, simulation length and number of conformations) and content (\textit{e.g.}, lipid compositions and ion concentrations) enable analyses of rare phenomena that are beyond the current possibilities for a single research group. Such analyses also pave the way for applications of MD simulations in new fields as demonstrated here by analysing the anisotropic diffusion of water in membrane systems (Fig.~\ref{fig:permeability}) that is essential parameter in pharmacokinetic modeling and MRI imaging~\cite{nitsche19,topgaard20}. Furthermore, the NMRlipids databank-API delivers access to a training set data that paves the way for diverse machine learning applications to 
%For example, a machine learning model predicting electron density profiles from form factors would be highly useful in interpretation of scattering experiments.
%In addition, the data contained in the NMRlipids databank will enable training of more elaborated models to
predict membrane properties. Such applications are analogous to AlphaFold \cite{jumper21} and other tools that predict proteins structures from their sequence using artificial intelligence. These possibilities are particularly valuable for scientist who do not typically have access to large scale MD simulation data. Different types of expected applications of the NMRlipids databank in wide range of fields are listed in Table~\ref{tab:applications}, yet the scope of such applications
%in biology, biotechnology, material science, biomolecular imaging and beyond
is expected to further widen with increasing amount of data in the NMRlipids databank.

\begin{table}[t]
    \centering
    \begin{tabular}{p{5.0cm}  p{5.0cm}  p{4.0cm}}
    Type of application     & Practical examples & Target group \\
    \hline
    Analyses of rare phenomena               & Lipid flip--flops, water permeation (Figs.~\ref{fig:flip-flops} and~\ref{fig:permeability}) & Membrane scientists \\
    \\
    Correlations between membrane properties & 
    Membrane structural properties, water dynamics (Figs.~\ref{fig:quality} and~\ref{fig:permeability}) & 
    Membrane scientists \\
    \\
    Applications that are outside typical scope of MD simulations & 
    Anisotropic water diffusion for pharmacokinetics and MRI imaging applications (Fig.~\ref{fig:permeability}) & 
    Scientists in fields where MD simulations are not usually applied \\
    \\
    Selection of the best simulation model for a specific application & 
    Best model for membranes with PC and PE lipids (Fig.~\ref{fig:quality}), lipid headgroup conformations~\cite{bacle21}, 
    packing of PS~\cite{antila22b} and PE (Fig.~\ref{fig:quality}) containing membranes. &
    Scientists using MD simulations \\
    \\
    Guidance for force field development & 
    Improvements in ion binding to lipids~\cite{melcr18,melcr20} and lipid headgroup conformational ensembles~\cite{yu21,dickson22,grote20} &
    Scientists developing parameters for MD simulations \\
    \\
    Training and target data for the development of coarse grained models & 
    Optimizing parameters of coarse grained models against NMRlipids databank, extracting continuum parameters for membranes. &
    Scientists developing and using coarse grained MD simulations \\
    \\
    Training set for machine learning applicatons &
    Programmatic access to the data and results enables training of machine learning models for various applications, such as predictions of membrane properties from composition & Scientists building and using machine learning applications for biomolecules.  
    \end{tabular}
    \caption{Examples on applications of the NMRlipids databank.}
    \label{tab:applications}
\end{table}

Potential benefits of sharing MD simulation data are widely recognized by different stakeholders from scientists~\cite{feig99,tai04,silva06,abraham19,hildebrand19,hospital20,abriata20,espigares20} to funders and journal staff, but data sharing with efficient tools for upcycling have been rare. The main barriers for such solutions have been the required commitment for the long term support of hardware and software, and the lack of incentives for researchers to share the data. The first issue is solved in the NMRlipids databank using the overlay design where raw data is distributed to already publicly available decentralized locations while the core of the databank is composed only of metadata stored in a version controlled git repository with an open access licence. On the other hand, the open collaboration approach developed in the NMRlipids project~\cite{botan15} creates incentives for sharing the data by offering authorship in published articles to the contributors. Advantages of such approach are demonstrated here for membrane simulations, but the concept can be applied also not only to other biomolecules, such as disordered proteins and membrane--protein systems, but also in other fields where similar barriers hinder the machine learning revolution, such as assignment of NMR spectra~\cite{klukowski22}.

@@

\subsection{Structure of the databank}
The overlay structure designed for the NMRlipids databank is composed of three layers illustrated in Fig.~\ref{fig:overlay}A. The {\it data layer} contains raw data that can be distributed to publicly available servers such as Zenodo (\url{www.zenodo.org}). The core content of the databank locates in the {\it databank layer} which is a git repository at \url{https://github.com/NMRLipids/Databank/} and is also permanently stored in Zenodo repository (\url{www.zenodo.org})~\cite{??}. The Essential information of each simulation is stored in a human and machine readable README.yaml file located at subfolders of \texttt{/Data/Simulations} folder in the {\it databank layer} repository. Each subfolder has a unique name constructed based on hash code of trajectory and topology files of each simulation. The README.yaml files in these folders contain access to all information that are needed for further analyses of simulations, such as links to the raw data and affiliations of universal molecule and atom names. The content of these files is described in detail in Table~\ref{tab:READMEkeys} in the supplementary information. Results from analyses of basic membrane properties (area per lipid, thickness, C-H bond order parameters, x-ray scattering form factors and relaxation of principal components) are stored in the same folders with README.yaml files. Experimental data used for ranking is stored in \texttt{/Data/experiments} folder, results from such ranking in \texttt{/Data/Ranking} and  relevant scripts in \texttt{/Scripts/} folder in the {\it databank layer} repository. 
%Raw MD simulation data are stored in external publicly available and stable repositories (\textit{Data layer} in Fig.~\ref{fig:overlay}), such as Zenodo (\url{www.zenodo.org}), from where it can be downloaded whenever needed using the links in README.yaml files.  
The scripts in the NMRlipids databank are mainly written in Python and many of them utilize the MDAnalysis module~\cite{gowers2019mdanalysis,michaud2011mdanalysis}. The databank structure is illustrated more detailed in Fig.~\ref{DatabankStructure} in the supplementary information. Whenever specific files or folders are referred here, they locate at {\it databank layer} repository unless stated otherwise.

\subsection{Molecule and atom naming convention} \label{naming}
When analysing simulations, atoms and molecules needs to be often called by their names defined in the simulation trajectory. However, these names typically vary between force fields because the universal naming convention has not been defined for lipids. To enable automatic analyses over simulations with different atom and molecule names in the NMRlipids databank, we have defined unique naming conventions for molecules and atoms therein. Unique abbreviations used in the NMRlipids databank for each molecule are listed in Table~\ref{tab:abbreviations} in the supplementary information. Atom names used in simulation trajectories are connected to unique atom names using mapping files that are defined in the NMRlipids project (\url{https://nmrlipids.blogspot.com/2022/04/new-yaml-format-of-mapping-files.html}). These files are located at \texttt{/Scripts/BuildDatabank/mapping\_files} in the NMRlipids databank repository. These files also define whether an atom belongs to headgroup, glycerol backbone, or acyl chain region in a lipid. In practise, force field specific molecule names and mapping files
names are given in the COMPOSITION dictionary in README.yaml files for each molecule in each simulation in the NMRlipids databank.

>>>>

Here we present the NMRlipids Databank --- a community-driven, open-for-all database featuring programmatic access to atom-resolution MD simulations of lipid bilayers. The programmatic access enables users to apply their data-driven approaches on the Databank, thus facilitating the creation of new tools for (and by) researchers in a wide range of fields covering academia and industry, from cell membrane biology and lipid nanoparticle formulations to computational chemistry and machine learning. As two usage examples of what is already possible, we demonstrate here (i) how data-driven analysis of water anisotropic diffusion in all the membrane systems available in the Databank can extend the scope of MD simulations to magnetic resonance imaging (MRI) and pharmacokinetics, and (ii) how the Databank allows its users to analyse rare phenomena that are beyond the scope of standard MD simulation investigations. Wider adaptation of the NMRlipids Databank will open even more possibilities. Furthermore, the Databank performs automatic quality evaluation of membrane simulations, which facilitates the selection of best-performing models for each given application and accelerates the development of simulation parameters and methodology.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\linewidth]{Figures/overlay3.pdf}
    \caption{A) Schematic presentation of the overlay structure used in the NMRlipids Databank. A more detailed structure of the \textit{Databank layer} is shown in Fig.~\ref{DatabankStructure} in the SI.
    %b) Distribution of the lengths of the trajectories, total number of trajectories and total length of the simulations in the NMRlipids databank.
    B) Distribution of lipids present in the trajectories of the Databank. `Others' lists lipids occurring in five or fewer simulations. 
    C) Distribution of force fields in the simulations in the Databank. References for each force field are given in Table~\ref{tab:ForceFields}.
    D) Flowchart for performing an analysis of properties through all MD simulations in the NMRlipids Databank using the API.
    E) Flowchart for accessing results calculated from the NMRlipids Databank and stored to the \textit{Application layer}.
    F) Currently available single lipid component and binary mixtures in the NMRlipids Databank. Colorbar shows the number of available simulations with the darkest green indicating three or more.  
    }
    \label{fig:overlay}
\end{figure}

@@

\begin{figure}[!t]
    \centering
    \includegraphics[width=\linewidth]{Figures/quality5.pdf}
    \caption{ A) Area per lipid of POPC and POPE lipid bilayers predicted by different force fields at 310\,K in simulations that are available in the NMRlipids Databank. The data points from the best-performing simulations, based on rankings in panels B and C, are surrounded by black circles. 
    B) Best POPC simulations ranked based on the \textit{sn}-1 acyl chain order parameter quality ($P^\mathrm{sn1}$). Also \textit{sn}-2 acyl chain ($P^\mathrm{sn2}$), headgroup ($P^\mathrm{hg}$) and total ($P^\mathrm{total}$) order parameter qualities, form factor quality ($FF_q$), and relative equilibration time for conformations ($\tau_\mathrm{rel}$) are shown.
    C) Best POPE simulations ranked based on the \textit{sn}-1 acyl chain order parameter quality. 
    D)--F) Direct comparison against experimental (NMR order parameters and X-ray scattering ) data exemplified for a simulation with the best overall order parameter quality (D), the best quality for POPE lipid (E), and the headgroup quality for POPE (F).
    G) Scatter plots and Pearson correlation coefficients, $r$, for the membrane area per lipid, thickness, first minimum of X-ray scattering form factor and average order parameter of the \textit{sn}-1 acyl chain extracted from the NMRlipids Databank. All correlation coefficients have p-value below 0.001. For more correlations see Fig.~\ref{fig:QualityCorrelationsSI}.
    }
    \label{fig:quality}
\end{figure}

@@

\begin{figure}[htb]
    \centering
    \includegraphics[width=88mm]{Figures/CholFlipFlops.pdf}
    \caption{\textbf{A} Illustration of cholesterol flip-flop.  
      \textbf{B--D} Cholesterol flip-flops analyzed from the Databank as a function of membrane thickness, area per lipid, and acyl chain order. Values from simulations with non-zero flip-flop rates are shown with blue dots. Histogrammed values are shown with black dots. Error bars show the standard error of the mean within each bin.
    }
    \label{fig:flip-flops}
\end{figure}

@@

\begin{figure}[tb]
    \centering
    \includegraphics[width=\linewidth]{Figures/permeation3.pdf}
    \caption{\textbf{A} Water diffusion, $D_\perp$, and permeability, $P$, through membranes, and lateral diffusion along the membrane, $D_\parallel$, illustrated in a multilamellar stack of lipid bilayers. 
    \textbf{B--E} Water permeation through membranes analyzed from the Databank as a function of temperature, thickness, area per lipid, and acyl chain order. Inset in B) shows the Arrhenius plot of permeation ($\ln(P)$ vs. $1/T$) that gives 17\,$\pm$\,4\,$k_\mathrm{B}T$ for the average activation energy for water permeation through lipid bilayer. Insets in C) and D) show the region where the dependence could be considered approximately linear.
    \textbf{F} Lateral diffusion of water as a function of hydration level. Experimental points for DMPC bilayers at 313\,K at different hydration levels are shown~\cite{rudakova04}.
    \textbf{G--H} Diffusion anistoropy of water as a function of thickness and area per lipid. 
    Non-zero permeation and diffusion values from simulations are shown with blue dots. Histogrammed values are shown with black dots. Error bars show the standard error of the mean within each bin. Only bins with more than one microsecond of data in total were used for water permeation.}
    \label{fig:permeability}
\end{figure}

To this end, we first calculated the water permeability through membranes from all simulations in the NMRlipids Databank using the general workflow depicted in Fig.~\ref{fig:overlay}D. The resulting non-zero values range between 0.3--322\,\textmu{}m/s with the mean of 14\,\textmu{}m/s and median of 8\,\textmu{}m/s. These values agree with the previously reported simulation results~\cite{venable19,camilo2022}, but are on average slightly larger than experimental values reported for PC lipids in the liquid crystalline phase,~0.19--0.33\,\textmu{}m/s~\cite{jansen95}. Using the workflow depicted in Fig.~\ref{fig:overlay}E, we then plotted the observed permeabilities and their histogrammed values in Figs.~\ref{fig:permeability}B--E as a function of temperature, membrane thickness, area per lipid, and acyl chain order. As expected, the permeability increases with the temperature, giving an average energy barrier of 17\,$\pm$\,4\,$k_\mathrm{B}T$ for the water permeation from the Arrhenius plot in Fig.~\ref{fig:permeability}B. On the other hand, the water permeability on average decreases when membranes become more packed, that is, with decreasing area per lipid and increasing thickness and acyl chain order (Figs.~\ref{fig:permeability}C--E). Permeation of water through bilayers depends on membrane properties also according to previous studies, but there is no established consensus on whether the area per lipid~\cite{nagle08} or bilayer thickness~\cite{frallicciardi22} is the main parameter determining the permeability. 
%is previously reported from simulations and experiments for some systems, but not for all~\cite{mathai07,venable19,frallicciardi22}. 
Our analysis over the NMRlipids Databank, containing significantly more data than what was available in previous studies, suggest non-linear dependencies on both of these parameters, yet the linear correlation would be a good approximation for thicknesses $\gtrsim$\,3.9\,nm and area per lipids $\lesssim$\,69\,\AA{}$^2$ (insets in Figs.~\ref{fig:permeability}C--D).
%within the range of experimentally reported values  permeability which is 1-2 orders of magnitude slower than osmotic permeability \cite{jansen95,??}. 
Clear dependencies of permeability on hydration level or the fraction of charged lipids, cholesterol, or POPE in the membrane were not observed (Fig.~\ref{fig:permeationSI}).
%
%%Several models on how water dynamics depends on membrane properties have been proposed~\cite{mathai07,nitsche13,nitsche16,shinoda16,venable19,frallicciardi22}.

To analyze how water diffusion anisotropy depends on membrane properties in a multi-lamellar lipid bilayer system, we calculated the water diffusion parallel to the membrane surface from all simulations in the NMRlipids Databank using the general workflow depicted in Fig.~\ref{fig:overlay}D. The parallel diffusion coefficient of water, $D_\parallel$, decreases with reduced hydration and increases with the temperature, but dependencies on the membrane area per lipid, thickness, or fraction of charged lipids were not observed in Figs.~\ref{fig:permeability} and~\ref{fig:diffusionSI} when the workflow depicted in Fig.~\ref{fig:overlay}E was used. Simulation results are close to the experimental values with low hydration levels in Fig.~\ref{fig:permeability}F, but increase to approximately two times higher than the experimental value for bulk water diffusion value (3.1$\times$\,10$^{-9}$\,m$^2$/s at 313\,K~\cite{khakimov08}) with high hydration levels.
This is not surprising as the most common water model used in membrane simulations, TIP3P, overestimates the bulk water diffusion~\cite{pathirannahalage21}. To estimate the diffusion anisotropy of water, $D_\perp/D_\parallel$, in multilamellar membrane system, the permeability coefficients of water through membranes were translated to perpendicular diffusion coefficients, $D_\perp$, using the Tanner equation~\cite{tanner78,wasterby02}. The resulting perpendicular diffusion coefficients are approximately five orders of magnitude smaller than the lateral diffusion coefficients of water (Figs.~\ref{fig:permeability}G--H), which is at the upper limit of anisotropy estimated from experimental data~\cite{nitsche19}. A significant increase in the diffusion anisotropy with membrane packing is observed, as $D_\perp/D_\parallel$ deviates further from unity with decreasing area per lipid and increasing thickness in Figs.~\ref{fig:permeability}G--H. This follows from decreasing water permeability with membrane packing (Figs.~\ref{fig:permeability}C--D), while lateral diffusion remains approximately constant (Figs.~\ref{fig:diffusionSI}A,~C).

In summary, our results suggest that the bilayer packing has a substantial effect on anisotropic water diffusion in multi-membrane lipid systems. The several-fold larger anisotropy in membranes with higher lateral density is expected to play a role in pharmacokinetic models not only for water but also for other hydrophilic molecules~\cite{nitsche19}. Furthermore, the enhanced understanding of this anisotropy may help in developing new diffusion-tensor-based MRI imaging methods where signals originate from the anisotropic diffusion of water in biological matter~\cite{topgaard20}.

@@

The focus of biomolecular simulations is moving from studies of individual molecules to larger complexes and even whole cells and organelles~\cite{johnson15,thornburg22,gupta22}. Simultaneously, machine-learning-based models for predicting the behaviour of biomolecules and automatic approaches to parametrize models are emerging~\cite{jumper21,antila22b}. The resources delivered by the NMRlipids Databank will support the development in both of these directions. Automatic quality evaluation and ranking of simulations against experimental data enable the selection of best simulations for specific applications without laborious manual force field evaluation. This also streamlines automatic parametrization procedures for atomistic and coarse grained simulations. Such practises for fostering the accuracy of simulations are becoming increasingly important due the accumulation of small errors when complexity and size of simulated systems are increasing.

The open programmatic access to unprecedented amounts of MD simulation data through the NMRlipids Databank-API enables a wide range of novel data-driven analyses and machine learning applications. The analyses of cholesterol flip-flop events (Fig.~\ref{fig:flip-flops}) and water permeation through membranes (Fig.~\ref{fig:permeability}) demonstrate how a large amount of accessible simulation data in terms of quantity (\textit{e.g.}, simulation length and number of conformations) and content (\textit{e.g.}, lipid compositions and ion concentrations) enable analyses of rare phenomena that are beyond the current possibilities for a single research group. Such analyses also pave the way for applications of MD simulations in new fields, as demonstrated here by analysing an essential parameter in pharmacokinetic modeling and MRI imaging:\cite{nitsche19,topgaard20} the anisotropic diffusion of water in membrane systems (Fig.~\ref{fig:permeability}). Furthermore, the NMRlipids Databank-API delivers access to a training set data that paves the way for diverse machine learning applications to 
%For example, a machine learning model predicting electron density profiles from form factors would be highly useful in interpretation of scattering experiments.
%In addition, the data contained in the NMRlipids databank will enable training of more elaborated models to
predict membrane properties. Such applications are analogous to AlphaFold \cite{jumper21} and other tools that predict protein structures from their sequence using artificial intelligence. These possibilities are particularly valuable for scientists who do not typically have access to large scale MD simulation data. Expected applications of the NMRlipids Databank in a wide range of fields are listed in Table~\ref{tab:applications}, yet the scope of such applications
%in biology, biotechnology, material science, biomolecular imaging and beyond
is expected to further widen with increasing amount of data in the Databank.

\begin{table}[t]
    \centering
    \begin{tabular}{p{5.0cm}  p{5.0cm}  p{4.0cm}}
    Type of application     & Practical examples & Target group \\
    \hline
    Analyses of rare phenomena               & Lipid flip-flops, water permeation (Figs.~\ref{fig:flip-flops} and~\ref{fig:permeability}) & Membrane scientists \\
    \\
    Correlations between membrane properties & 
    Membrane structural properties, water dynamics (Figs.~\ref{fig:quality} and~\ref{fig:permeability}) & 
    Membrane scientists \\
    \\
    Applications that are outside the typical scope of MD simulations & 
    Anisotropic water diffusion for pharmacokinetics and MRI imaging applications (Fig.~\ref{fig:permeability}) & 
    Scientists in fields where MD simulations are not usually applied \\
    \\
    Selection of the best simulation model for a specific application & 
    Best model for membranes with PC and PE lipids (Fig.~\ref{fig:quality}), lipid headgroup conformations~\cite{bacle21}, 
    packing of PS~\cite{antila22b} and PE (Fig.~\ref{fig:quality}) containing membranes &
    Scientists using MD simulations \\
    \\
    Guidance for force field development & 
    Improvements in ion binding to lipids~\cite{melcr18,melcr20} and lipid headgroup conformational ensembles~\cite{yu21,dickson22,grote20} &
    Scientists developing parameters for MD simulations \\
    \\
    Training and target data for the development of coarse grained models & 
    Optimizing parameters of coarse grained models against NMRlipids Databank, extracting continuum parameters for membranes &
    Scientists developing and using coarse grained MD simulations \\
    \\
    Training set for machine learning applicatons &
    Programmatic access to the data and results enables training of machine learning models for various applications, such as predictions of membrane properties from composition & Scientists building and using machine learning applications for biomolecules
    \end{tabular}
    \caption{Examples on applications of the NMRlipids Databank.}
    \label{tab:applications}
\end{table}

Potential benefits of sharing MD simulation data are widely recognized by different stakeholders from scientists~\cite{feig99,tai04,silva06,abraham19,hildebrand19,hospital20,abriata20,espigares20} to funders and journal staff, but data sharing with efficient tools for upcycling have been rare. The main barriers for such solutions have been the required commitment for the long term support of hardware and software, and the lack of incentives for researchers to share the data. The first issue is solved in the NMRlipids Databank by the overlay design, where the raw data is distributed to already publicly available decentralized locations, while the core of the Databank is composed only of the metadata stored in a version-controlled git repository with an open-access license. On the other hand, the open collaboration approach developed in the NMRlipids Project~\cite{botan15} creates incentives for sharing the data by offering authorship in published articles to the contributors. Advantages of such an approach are demonstrated here for membrane simulations, but the concept can be applied not only to other biomolecules, such as disordered proteins and membrane--protein systems, but also in other fields where similar barriers hinder the machine learning revolution, such as the assignment of NMR spectra~\cite{klukowski22}.

@@

\subsection{Structure of the Databank}
The overlay structure designed for the NMRlipids Databank is composed of three layers (Fig.~\ref{fig:overlay}A).
The {\it Data layer} contains raw data that can be distributed to publicly available servers such as Zenodo (\href{https://www.zenodo.org/}{zenodo.org}).
The core content of the Databank locates in the {\it Databank layer}, which is a git repository at \href{https://github.com/NMRlipids/Databank/}{github.com/NMRlipids/Databank} and is also permanently stored in a Zenodo repository (\href{https://www.zenodo.org/}{zenodo.org})~\cite{??}. The essential information of each simulation is stored in a human-and-machine-readable README.yaml file located in a subfolder of the \texttt{/Data/Simulations} folder in the {\it Databank layer} repository; each subfolder has a unique name constructed based on a hash code of the trajectory and topology files of each simulation. The README.yaml files in these folders contain access to all information that is needed for further analysis of simulations, such as links to the raw data and associations with the universal molecule and atom names. The content of these files is described in detail in Table~\ref{tab:READMEkeys} in the supplementary information. Results from analyses of basic membrane properties (area per lipid, thickness, C--H bond order parameters, X-ray scattering form factors, and relaxation of principal components) are stored in the same folders as the README.yaml files.
Experimental data used for ranking is stored in the \texttt{/Data/experiments} folder,
the ranking results in \texttt{/Data/Ranking},
and the relevant scripts in the \texttt{/Scripts/} folder in the {\it Databank layer} repository. 
%Raw MD simulation data are stored in external publicly available and stable repositories (\textit{Data layer} in Fig.~\ref{fig:overlay}), such as Zenodo (\url{www.zenodo.org}), from where it can be downloaded whenever needed using the links in README.yaml files.  
The scripts in the NMRlipids Databank are mainly written in Python and many of them use the MDAnalysis module~\cite{gowers2019mdanalysis,michaud2011mdanalysis}. The Databank structure is illustrated in more detail in Fig.~\ref{DatabankStructure} in the supplementary information. Whenever specific files or folders are referred here, they locate at the {\it Databank layer} repository unless stated otherwise.

\subsection{Universal naming convention for molecules and atoms} \label{naming}
When analysing simulation trajectories, atoms and molecules often need to be  called by the names used in the trajectory. However, these names typically vary between force fields, as a universal naming convention has not been defined for lipids. To enable automatic analyses over all the simulations in the NMRlipids Databank, we have defined universal naming conventions for the molecules and atoms therein. The universal abbreviations used in the NMRlipids Databank for each molecule are listed in Table~\ref{tab:abbreviations} in the supplementary information. The atom names used in simulation trajectories are connected to the universal atom names using the mapping files defined in the NMRlipids Project (\url{https://nmrlipids.blogspot.com/2022/04/new-yaml-format-of-mapping-files.html}). These files are located at \texttt{/Scripts/BuildDatabank/mapping\_files} in the NMRlipids Databank repository. These files also define whether an atom belongs to the headgroup, glycerol backbone, or acyl chain region in a lipid. In practise, the force-field-specific molecule names and mapping file
names are given in the COMPOSITION dictionary in the README.yaml files for each molecule in each simulation in the NMRlipids Databank.
